---
title: "Exploring diabetes in the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---
__Dataset:__ Behavioral Risk Factor Surveillance System Data <br><br>
__Source:__ https://www.cdc.gov/brfss/<br><br>
__Summary:__ The data set is a collection of responses from ~400,000 questionnaires across the United States on health related topics. BRFSS started in the 1980s and has been used to monitor pervasive illnesses in the US through questionnaires. The research is a standardized questionairre with additional questions varying by state. Interviewers reach partipant in house and by telephone. The telephone surveys are conducted using "Random Digit Dialing (RDD) techniques on both landlines and cell phones" so much of the data appears to be from random sampling. The research is retroactive and not a designed experiment so while correlation can be inferred, causality cannot.

The features in the data set are both continuous and categorical. Some features are missing values so data preprocessing is essential. Because the participants were selected randomly the sample can be generalized. <br><br>
__Goals:__ Explore any correlations among gender, weight, and age<br><br>

## Part 0: Setup
```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(Rgraphviz)
library(knitr)
library(grid)
library(gridExtra)
```

```{r load-data}
load("brfss2013.RData")
```

```{r}
# group and count a feature with discrete values
feature_vcounts <- function(df, f) {
  df %>%
    group_by_at(f) %>%
      count()}

# method for binning values
bin_min_sample <- function(p) {
  n = 10
  a = 10/p
  b = 10/(1-p)
  max(c(a,b))}

# create a new df for simulating binom probability distribution
binom_prob_df <- function(df, f, target) {
  new_df <- feature_vcounts(df,f)
  new_df$n[new_df[f] == target]/sum(new_df$n)}

# filtering df with subgroup value
subgroup_df <- function(df,f, group) {
  filter(df,df[f]==group)}

# calc the vector probability
binom_prob_vec <- function(v, target) {
  sum(v == target)/length(v)}

# sample from df
binom_sample <- function(s,v)
  sample(v, size=s, replace=TRUE)

# create the binomial sample distribution
binom_sample_dist <- function(df,f,target) {
  sample_dist <- c()
  for (i in 1:10001) {
    prob <- binom_prob_vec(binom_sample(100,df[,f]),target)
    sample_dist <- append(sample_dist,prob)}
  return(sample_dist)}

# convert decimal to percent
to_percent <- function(pvalue) {
  paste(round(pvalue*100,digits= 2),"%",sep="")}
```

* * *

## Part 1: Data

Import and filter the data to only include important features relevant to diabetes, gender, weight, and age. The features are reduced 4 columns and then the data is cleaned to remove rows that have invalid data or data that has less pertinent information.
```{r error=FALSE, warning=FALSE,results='asis'}
# Import original file:
orig_dim <- dim(brfss2013)

# Select only relevant features:
weight_diabetes <- brfss2013 %>%
   select(sex, X_ageg5yr, weight2,diabete3)

# ------------------Cleaning data------------------
# 1.Weight strings -> numeric
weight_diabetes$weight2 <- as.numeric(as.character(weight_diabetes$weight2))
new_dim <- dim(weight_diabetes)

# 2. Remove Null Weights and Weights over 400
weight_diabetes <- na.omit(weight_diabetes)
weight_diabetes <- filter(weight_diabetes, weight2 <= 400)

# 3. Remove Diabetes Responses
target <- c("Yes", "No")
weight_diabetes <- filter(weight_diabetes, diabete3 %in% target)

# 4. Add index and reorder
weight_diabetes$index <- seq.int(nrow(weight_diabetes))
weight_diabetes <- weight_diabetes[c(4,3,1,2)]
clean_dim <- dim(weight_diabetes)

# Show data:
kable(head(weight_diabetes,n=5), caption="Diabetes Data Set",padding=0, format = "markdown",align="l")
```
The data looks simplified and only contains the features wanted for this project. Because the data needs to be anonymized, age ranges are a safe alternative to specific ages. Age ranges will be used as categorial information for this data set.

__Original dimensions:__ <br>[`r orig_dim[1]`] x [`r orig_dim[2]`]<br>
__Reduced dimensions:__ <br> [`r new_dim[1]`] x [`r new_dim[2]`]<br>
__After cleaning:__ <br>[`r clean_dim[1]`] x [`r clean_dim[2]`]

<br>

* * *

## Part 2: Research questions

### **Research quesion 1:**<br>
##### Is there any correlation among sex, weight, and age? (vars: sex, weight2, X_ageg5yr)<br>
Because gender is a key variable in biometrics, it is important to explore whether gender might correllate to other variables. In this case, we are looking at whether or not sex correllates with weight.<br>

### **Research question 2:**<br>
#### Do weight, age, and/or gender correlate with diabetes? How? (vars: sex, X_ageg5yr, weight2, diabete3)<br>
The goal of this exploratory project is to determine whether or not weight/gender correlate with diabetes. Knowledge of any correlation could be helpful with regards to informing patients of their likelihood of having diabetes based on their gender and weight.<br>


### **Research question 3:**<br>
#### What is the liklihood that a participant from the sample reports having diabetes given that they are male and weigh 200 lbs or more (given the sample data)? (vars: sex, weight2, diabete3)<br>
Knowing the liklihood of having a positive diagnosis for diabetes can be valuable for both researchers and patients. We will look at the example scenario and determine what the likelihood is for a male who weighs more than 200 lbs to report having diabetes. 
<br>

* * *

## Part 3: Exploratory data analysis

### **Research question 1:**
##### Is there any correlation among sex, weight, and age? (vars: sex, weight2, X_ageg5yr)<br>
_Weight2_ is continuous  so it is important to first check the distribution of the data. _Sex_ is binary categorical so we will visualize it's distribution with a bar plot.<br>

```{r fig.width=12, fig.height=3, error=FALSE, warning=FALSE, message=FALSE}
centered <- theme(plot.title = element_text(hjust = 0.5))
hist_weight <- ggplot(data=weight_diabetes,aes(weight2,  fill=weight2))+
  geom_histogram(fill='salmon',color='white') + ggtitle("Histogram [Weight]") + centered
weight_diabetes$log_weight <- log(weight_diabetes$weight2)
hist_log_weight <- ggplot(data=weight_diabetes,aes(log_weight, fill=log_weight))+
  geom_histogram(fill='mediumturquoise',color='white') + ggtitle("Histogram [Log_Weight]") + centered
grid.arrange(hist_weight, hist_log_weight, ncol = 2)
```
<br>For _Weight2_, the distribution is right skewed while the log of _Weight2_ is nearly normal. Because the data is nearly normal and unimodal, a frequentist approach using inferential statistics can be used.
<br>

```{r fig.width=4, fig.height=3, error=FALSE, warning=FALSE, message=FALSE}

gender_count <- ggplot(data=weight_diabetes,aes(sex, fill = sex))+
  geom_bar(color='white') + ggtitle("Gender") + centered 
gender_count
```
<br>There are more female participants than male participants by a margin much larger than the general population in the US. This could be an indicator that the sampling method was not entirely random with regards to gender sampling. Nonetheles, the data sample is large enough to continue evaluating health risk factors.

<br>
```{r fig.width=12, fig.height=3, error=FALSE, warning=FALSE, message=FALSE}
age_count <- ggplot(data=weight_diabetes,aes(X_ageg5yr)) +
  geom_bar(fill='salmon',color='white') + ggtitle("Age") + centered
age_count
```
The age range appears to be left skewed with extremes at both tails. The large tails are most likely a result of binning where the youngest and oldest age groups account for larger age ranges.
<br>

```{r fig.width=12, fig.height=3, error=FALSE, warning=FALSE, message=FALSE}
ggplot(weight_diabetes, aes(x=X_ageg5yr,y=weight2, color=sex)) +
  geom_boxplot()
```
When comparing age to weight, there does seem to be a clear difference in weigh distributions for genders. Male's appear to weigh more than females. <br><br>It should also be noted that there also appear to be adult patients who weigh less than 50 lbs. Future analysis of the data collection process should explore whether these low weight outliers are mistakes or if they reflect patients with serious health issues.


### **Research question 2:**
#### Do weight, age, and/or gender correlate with diabetes? How? (vars: sex, X_ageg5yr, weight2, diabete3)<br>
```{r fig.width=12, fig.height=3, error=FALSE, warning=FALSE, message=FALSE}
ggplot(weight_diabetes, aes(x=X_ageg5yr,y=weight2, color=diabete3)) +
  geom_boxplot() + 
  scale_y_continuous()

ggplot(weight_diabetes, aes(x = factor(X_ageg5yr), fill = factor(diabete3))) +
  geom_bar(position="fill")
```
```{r fig.width=4, fig.height=3,}
ggplot(weight_diabetes, aes(x = factor(sex), fill = factor(diabete3))) +
  geom_bar(position="fill")
```
<br>When looking at female and male participants in the sample, the ratio of reporting diabetes appears is very similar. It is likely that difference in gender does not correllate to reporting of diabetes.

### **Research question 3:**
#### What is the liklihood that a participant from the sample reports having diabetes given that they are male and weigh 200 lbs or more (given the sample data)? (vars: sex, weight2, diabete3)<br>
```{r fig.width=12, fig.height=3, warning=FALSE}
sex_male_distribution <- binom_sample_dist(weight_diabetes,"sex","Male")

male_plot<-ggplot()+
  aes(x=sex_male_distribution)+
  geom_bar(fill='salmon',color='white') + ggtitle("Female - Binom Distribution") + centered

gp_male <- binom_prob_vec(weight_diabetes$sex,"Male")

sex_female_distribution <- binom_sample_dist(weight_diabetes,"sex","Female")

female_plot<-ggplot()+
  aes(x=sex_female_distribution)+
  geom_bar(fill='mediumturquoise',color='white') + ggtitle("Male - Binomial Distribution") + centered

gp_female <- binom_prob_vec(weight_diabetes$sex,"Female")


grid.arrange(male_plot, female_plot, ncol = 2)
```
<br> For _Sex_, the probability of the participant being male is __`r to_percent(gp_male)`__ <br>
<br> The probability of the participant being female is __`r to_percent(gp_female)`__ <br>

```{r fig.width=12, fig.height=3, warning=FALSE}



diabetes_yes_distribution <- binom_sample_dist(weight_diabetes,"diabete3","Yes")
yes_plot<- ggplot()+
  aes(x=diabetes_yes_distribution)+
  geom_bar(fill='salmon',color='white') + ggtitle("Yes - Binom Dist") + centered
dp_yes <- binom_prob_vec(weight_diabetes$diabete3,"Yes")


diabetes_no_distribution <- binom_sample_dist(weight_diabetes,"diabete3","No")
no_plot<- ggplot()+
  aes(x=diabetes_no_distribution)+
  geom_bar(fill='mediumturquoise',color='white') + ggtitle("No - Binom Dist") + centered
dp_no <- binom_prob_vec(weight_diabetes$diabete3,"No")


grid.arrange(yes_plot, no_plot, ncol = 2)
```

<br>The probability of the participant reporting diabetes is low at around __`r to_percent(dp_yes)`__<br><br>
And the probability that a participant will report that they do not have diabetes is around __`r to_percent(dp_no)`__<br>

```{r warning=FALSE}
# probability that participant is 200 lbs or more using right tailed distribution
sd_weight <- sd(weight_diabetes$weight2)
mean_weight <- mean(weight_diabetes$weight2)
z <- (200-mean_weight)/sd_weight
righttailed <-1-pnorm(z)

# probability that participant is 200 lbs or more using lognormed distribution
sd_weight <- sd(weight_diabetes$log_weight)
mean_weight <- mean(weight_diabetes$log_weight)
z <- (log(200)-mean_weight)/sd_weight
1-pnorm(z)

# probability that participant is 200 lbs or more
pvalue=1-exp(pnorm(log(200),mean_weight,sd_weight, log=TRUE))
pvalue

# percentage of participants who weigh 200 or more lbs vs those who are under 200 lbs
weight_diabetes$heavy <- ifelse(weight_diabetes$weight2 < 200,"not_heavy","heavy")
actual <- binom_prob_vec(weight_diabetes$heavy,"heavy")
```
<br>
The likelihood that a randomly selected male from the sample reports having diabetes is  __`r to_percent(pvalue)`__. The right-tailed z-score estimate is __`r to_percent(righttailed)`__. The actual percentage is __`r to_percent(actual)`__ 
<br>

```{r}
#gp_male
#gp_female

#dp_no
#dp_yes

male_diabetes_prob <- binom_prob_vec(subgroup_df(weight_diabetes,'sex','Male')$diabete3,"Yes")
female_diabetes_prob <- binom_prob_vec(subgroup_df(weight_diabetes,'sex','Female')$diabete3,"Yes")

male_diabetes_prob
female_diabetes_prob
```

The probability that the participant reports having diabetes given that he is male is the probability that the participant is male multiplied by the probability that a participant reported having diabetes:


```{r}
#to_percent(gp_male * dp_yes)
```
The probability that a participant reported diabetes given that the participant is male is __`r to_percent(gp_male * dp_yes)`__.

```{r}
# Tutorial by: http://www.harrysurden.com/wordpress/archives/292
# R Conditional Probability Tree Diagram
 
# The Rgraphviz graphing package must be installed to do this
require("Rgraphviz")
 
# Assign probabilities
a<-round(gp_male,digits= 3)
bGivena<-round(male_diabetes_prob,digits= 3)
bGivenNota<-round(female_diabetes_prob,digits= 3)
 
# Calculations
notbGivena<-1-bGivena
notA<-1-a
notbGivenNota<-1-bGivenNota
aANDb<-a*bGivena
aANDnotb<-a*notbGivena
notaANDb <- notA*bGivenNota
notaANDnotb <- notA*notbGivenNota
b<- aANDb + notaANDb
notB <- 1-b
aGivenb <- aANDb / b

# Text notation 
node1<-"Pa"
node2<-"M"
node3<-"F"
node4<-"M&D"
node5<-"M&D'"
node6<-"F&D"
node7<-"F&D'"
nodeNames<-c(node1,node2,node3,node4, node5,node6, node7)
rEG <- new("graphNEL", nodes=nodeNames, edgemode="directed")
 
# Draw the "lines" or "branches" of the probability Tree
rEG <- addEdge(nodeNames[1], nodeNames[2], rEG, 1)
rEG <- addEdge(nodeNames[1], nodeNames[3], rEG, 1)
rEG <- addEdge(nodeNames[2], nodeNames[4], rEG, 1)
rEG <- addEdge(nodeNames[2], nodeNames[5], rEG, 1)
rEG <- addEdge(nodeNames[3], nodeNames[6], rEG, 1)
rEG <- addEdge(nodeNames[3], nodeNames[7], rEG, 10)
 
eAttrs <- list()
q<-edgeNames(rEG)

# Add the probability values to the the branch lines
eAttrs$label <- c(toString(a),toString(notA),
 toString(bGivena), toString(notbGivena),
 toString(bGivenNota), toString(notbGivenNota))
names(eAttrs$label) <- c(q[1],q[2], q[3], q[4], q[5], q[6])
edgeAttrs<-eAttrs
 
# Set the color, etc, of the tree
attributes<-list(node=list(label="foo", fillcolor="lightgreen", fontsize="15"),
 edge=list(color="red"),graph=list(rankdir="LR"))
 
#Plot the probability tree using Rgraphvis
plot(rEG, edgeAttrs=eAttrs, attrs=attributes)
#nodes(rEG)
#edges(rEG)
 
#Add the probability values to the leaves of A&B, A&B', A'&B, A'&B'
text(520,250,aANDb, cex=.8)
text(520,180,aANDnotb,cex=.8)
text(520,100,notaANDb,cex=.8)
text(520,30,notaANDnotb,cex=.8)
text(340,260,"(D | M)",cex=.8)
text(340,130,"(D | F')",cex=.8)
 
#Write a table in the lower left of the probablites of A and B
text(80,50,paste("P(M):",a),cex=.9, col="darkgreen")
text(80,20,paste("P(F):",notA),cex=.9, col="darkgreen")
text(160,50,paste("P(D):",round(b,digits=2)),cex=.9)
text(160,20,paste("P(D'):",round(notB, 2)),cex=.9)
text(100,220,paste("P(M|D): ",round(aGivenb,digits=2)),cex=.9,col="blue")
```
