---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

Import and filter the data to only include important features.

```{r}
orig_dim <- dim(brfss2013)
weight_diabetes <- brfss2013 %>%
   select(sex, weight2,diabete3)
new_dim <- dim(weight_diabetes)

sprintf("Original Dimensions: [%s]x[%s]  New Dimensions: [%s]x[%s]", orig_dim[1],orig_dim[2],new_dim[1],new_dim[2])
```
Vector has been reduced to 3 dimensions, but it still needs to be cleaned.

```{r}
head(weight_diabetes, n=10)
```

* * *

## Part 2: Research questions

**Research quesion 1:**<br>
What is the probability of having diabetes* based on gender?<br><br>
**Research quesion 2:**<br>
What is the distribution of weight from this data sample? What is the range of weights that fall within 66% of the sample set.<br><br>
**Research quesion 3:**<br>
Given results from RQ #2, what is the probability that a male over 200lbs has diabetes*?<br>

__Assumption:__ _The feature 'diabete3' is a survey question asking the participant whether or not they know they have diabetes. I will be interpreting the response of this survey question to be either a positive or negative diagnosis on diabetes, but in actuality data will reflect the respondants' awareness of whether or not they have diabetes._

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

```{r}

```



**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

